---
title: "Advanced and programmatic use"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced and programmatic use}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

This vignette is for users who want more control than the high-level wrappers
alone provide.

Topics covered:

- calling step functions directly
- understanding `ray_result`
- using selective steps and resume
- path helpers and input-mode helpers
- practical batch-style workflows

For most users, the main entry points remain:

- `ray_pipeline_single_tree()`
- `ray_pipeline_forest()`

# Images and Docker setup

Set the Docker image option before running pipelines.

```{r eval = FALSE}
library(rayforestr)

options(
  rayforestr.raycloud_image = "ghcr.io/csiro-robotics/raycloudtools:latest"
)

ray_check_docker()
```

In many setups, the same image also contains TreeTools commands such as
`treesmooth`, `treeinfo`, and `treemesh`.

# The `ray_result` object

Most step functions and wrappers return a `ray_result` object.

Typical fields include:

- `step`
- `status`
- `input`
- `outputs`
- `log_file`
- `details`

Example:

```{r eval = FALSE}
res <- ray_pipeline_single_tree(
  input_file   = "D:/packageR/data/single_tree/spruce_4_raycloud.ply",
  input_mode   = "raycloud",
  project_root = "D:/packageR",
  run_name     = "programmatic_demo",
  steps        = c("terrain", "trees"),
  resume       = FALSE,
  dry_run      = FALSE,
  echo         = TRUE
)

res
res$status
res$outputs
```

`ray_result` makes it easier to continue programmatically from later stages.

# Calling step functions directly

Advanced users may prefer to control each stage manually.

## Terrain

```{r eval = FALSE}
s2 <- ray_terrain(
  raycloud_file = "D:/packageR/data/single_tree/spruce_4_raycloud.ply",
  project_root  = "D:/packageR",
  out_dir       = "D:/packageR/runs/stepwise_demo/step4_terrain",
  dry_run       = FALSE,
  echo          = TRUE
)

s2$outputs$terrain_mesh
```

## Trees

```{r eval = FALSE}
s3 <- ray_trees(
  raycloud_file = "D:/packageR/data/single_tree/spruce_4_raycloud.ply",
  terrain_mesh  = s2$outputs$terrain_mesh,
  project_root  = "D:/packageR",
  out_dir       = "D:/packageR/runs/stepwise_demo/step5_trees",
  args_trees    = "--height_min 0.2 --distance_limit 3 --max_diameter 3 --gravity_factor 0.1",
  dry_run       = FALSE,
  echo          = TRUE
)

s3$outputs$treefile
```

## Smooth

```{r eval = FALSE}
s4 <- ray_smooth(
  treefile     = s3$outputs$treefile,
  project_root = "D:/packageR",
  out_dir      = "D:/packageR/runs/stepwise_demo/step7_smooth",
  dry_run      = FALSE,
  echo         = TRUE
)
```

## Info

```{r eval = FALSE}
s5 <- ray_info(
  treefile     = s4$outputs$smoothed_treefile,
  project_root = "D:/packageR",
  out_dir      = "D:/packageR/runs/stepwise_demo/step8_info",
  dry_run      = FALSE,
  echo         = TRUE
)
```

## Mesh

```{r eval = FALSE}
s6 <- ray_mesh(
  treefile     = s4$outputs$smoothed_treefile,
  project_root = "D:/packageR",
  out_dir      = "D:/packageR/runs/stepwise_demo/step9_mesh",
  dry_run      = FALSE,
  echo         = TRUE
)
```

# Selective steps in the wrappers

The wrappers support selective execution through `steps`.

```{r eval = FALSE}
res <- ray_pipeline_single_tree(
  input_file   = "D:/packageR/data/single_tree/spruce_4_raycloud.ply",
  input_mode   = "raycloud",
  project_root = "D:/packageR",
  run_name     = "selective_demo",
  steps        = c("terrain", "trees", "smooth"),
  resume       = FALSE,
  dry_run      = FALSE,
  echo         = TRUE
)
```

This is useful when you want only part of the full workflow.

# Resume behavior

A common advanced pattern is to split processing into multiple runs.

## First run

```{r eval = FALSE}
res_part1 <- ray_pipeline_single_tree(
  input_file   = "D:/packageR/data/single_tree/spruce_4_raycloud.ply",
  input_mode   = "raycloud",
  project_root = "D:/packageR",
  run_name     = "resume_demo",
  steps        = c("terrain", "trees"),
  resume       = FALSE,
  dry_run      = FALSE,
  echo         = TRUE
)
```

## Second run

```{r eval = FALSE}
res_part2 <- ray_pipeline_single_tree(
  input_file   = "D:/packageR/data/single_tree/spruce_4_raycloud.ply",
  input_mode   = "raycloud",
  project_root = "D:/packageR",
  run_name     = "resume_demo",
  steps        = c("smooth", "info", "mesh"),
  resume       = TRUE,
  dry_run      = FALSE,
  echo         = TRUE
)
```

The second call reuses the existing stage outputs in the same run directory.

# Input modes programmatically

The wrappers support multiple input modes.

```{r eval = FALSE}
ray_resolve_input_mode(NULL, "x/spruce_4.ply")
ray_resolve_input_mode(NULL, "x/spruce_4_raycloud.ply")
ray_resolve_input_mode(NULL, "x/spruce_4_raycloud_trees.txt")
ray_resolve_input_mode(NULL, "x/spruce_4_raycloud_trees_smoothed.txt")
```

Typical choices are:

- `point_cloud`
- `raycloud`
- `treefile`
- `smoothed_treefile`

# Path helpers

`rayforestr` includes helpers for creating consistent run folders.

```{r eval = FALSE}
paths <- ray_resolve_paths(
  input_file   = "D:/packageR/data/single_tree/spruce_4.ply",
  project_root = "D:/packageR",
  run_name     = "path_demo"
)

paths$out_root
paths$dirs
```

This is useful for advanced scripts or custom orchestration.

# Step metadata helpers

For programmatic workflows, it can be useful to inspect helper metadata.

```{r eval = FALSE}
ray_step_inputs()
ray_step_input_labels()
ray_result_statuses()
```

These can help when building custom wrappers or batch summaries.

# Batch-style workflow pattern

For many files, a simple pattern is:

1. create a job table
2. iterate over rows
3. call a wrapper
4. store result objects

Example:

```{r eval = FALSE}
files <- c(
  "D:/packageR/data/single_tree/tree1_raycloud.ply",
  "D:/packageR/data/single_tree/tree2_raycloud.ply"
)

results <- lapply(seq_along(files), function(i) {
  ray_pipeline_single_tree(
    input_file   = files[[i]],
    input_mode   = "raycloud",
    project_root = "D:/packageR",
    run_name     = paste0("batch_tree_", i),
    steps        = c("terrain", "trees", "smooth", "info", "mesh"),
    resume       = FALSE,
    dry_run      = FALSE,
    echo         = FALSE
  )
})

results
```

A similar pattern works for `ray_pipeline_forest()`.

# Reading outputs downstream

Programmatic use often continues with readers and summary functions.

```{r eval = FALSE}
tree_df <- ray_read_treefile(
  "D:/packageR/runs/example/step5_trees/spruce_4_raycloud_trees.txt"
)

info_df <- ray_read_treeinfo(
  "D:/packageR/runs/example/step8_info/spruce_4_raycloud_trees_smoothed_info.txt"
)
```

You can then use metric and volume helpers on the returned data.

# Practical tips

- Use `input_mode = "raycloud"` when a valid raycloud already exists.
- Use `steps` to avoid unnecessary reruns.
- Use `resume = TRUE` for multi-stage workflows.
- Keep one run name per logical analysis so stage reuse remains clear.
- Inspect `res$outputs` instead of reconstructing output paths manually.

# Notes

- Step functions are useful for advanced control and debugging.
- Wrappers are still recommended for most routine analyses.
- The stage-based design makes it easier to pause, inspect, and continue
  workflows without relying on copy-heavy pipeline logic.
