---
title: "Forest and plot workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forest and plot workflow}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`rayforestr` supports forest- and plot-level processing through
`ray_pipeline_forest()`. The wrapper organizes outputs into stage folders and
supports selective steps, resume, and multiple entry modes.

Typical forest/plot stages are:

- `import`
- `colour`
- `terrain`
- `trees`
- `split`
- `smooth`
- `info`
- `mesh`

This vignette shows recommended ways to use the forest wrapper.

# Prerequisites

Before running the forest workflow:

1. Install Docker and confirm that Docker is available from R.
2. Set the main image option.
3. Choose a project root outside the package source folder.

```{r eval = FALSE}
library(rayforestr)

ray_check_docker()

options(
  rayforestr.raycloud_image = "ghcr.io/csiro-robotics/raycloudtools:latest"
)
```

In many setups, the same image also contains `treesmooth`, `treeinfo`, and
`treemesh`, so no separate TreeTools image is required.

# Recommended project structure

```text
project_root/
  data/
    plot/
      plot01.ply
      plot01_raycloud.ply
  runs/
```

# Forest workflow from an existing raycloud

For many workflows, starting from an existing raycloud is the most practical
entry point.

```{r eval = FALSE}
res <- ray_pipeline_forest(
  input_file   = "D:/packageR/data/plot/plot01_raycloud.ply",
  input_mode   = "raycloud",
  project_root = "D:/packageR",
  run_name     = "forest_from_raycloud",
  use_colour   = FALSE,
  steps        = c("terrain", "trees", "split", "smooth", "info", "mesh"),
  resume       = FALSE,
  dry_run      = FALSE,
  echo         = TRUE
)

res
```

Expected stage folders include:

- `step4_terrain`
- `step5_trees`
- `step6_split`
- `step7_smooth`
- `step8_info`
- `step9_mesh`

# Forest workflow from a raw point cloud

```{r eval = FALSE}
res <- ray_pipeline_forest(
  input_file   = "D:/packageR/data/plot/plot01.ply",
  input_mode   = "point_cloud",
  project_root = "D:/packageR",
  run_name     = "forest_from_point_cloud",
  use_colour   = FALSE,
  steps        = c("import", "terrain", "trees", "split"),
  resume       = FALSE,
  dry_run      = FALSE,
  echo         = TRUE
)
```

## Import reuse modes

For raw point clouds, the wrapper supports the same import reuse logic as the
single-tree workflow:

- `reuse_existing_import = "never"`
- `reuse_existing_import = "prefer"`
- `reuse_existing_import = "fallback"`

### Example: fallback to a sibling raycloud

```{r eval = FALSE}
res <- ray_pipeline_forest(
  input_file = "D:/packageR/data/plot/plot01.ply",
  input_mode = "point_cloud",
  project_root = "D:/packageR",
  run_name = "forest_import_fallback",
  reuse_existing_import = "fallback",
  search_sibling_raycloud = TRUE,
  use_colour = FALSE,
  steps = c("import", "terrain", "trees", "split"),
  resume = FALSE,
  dry_run = FALSE,
  echo = TRUE
)
```

This is useful when raw import is noisy or unreliable but a valid
`plot01_raycloud.ply` already exists.

# Understanding the split stage

The `split` stage typically separates the segmented cloud into multiple PLY
files, for example by tree identifier or segment label.

The forest wrapper returns both:

- `split_segmented`
- `split_segmented_dir`

The directory path is usually the most useful output for downstream inspection.

```{r eval = FALSE}
res$outputs$split_segmented_dir
list.files(res$outputs$split_segmented_dir, full.names = TRUE)
```

# Resume and selective steps

A common workflow is to run early processing first, inspect results, and then
continue later.

## First run: terrain, trees, split

```{r eval = FALSE}
res_part1 <- ray_pipeline_forest(
  input_file   = "D:/packageR/data/plot/plot01_raycloud.ply",
  input_mode   = "raycloud",
  project_root = "D:/packageR",
  run_name     = "forest_resume_demo",
  use_colour   = FALSE,
  steps        = c("terrain", "trees", "split"),
  resume       = FALSE,
  dry_run      = FALSE,
  echo         = TRUE
)
```

## Second run: smooth, info, mesh

```{r eval = FALSE}
res_part2 <- ray_pipeline_forest(
  input_file   = "D:/packageR/data/plot/plot01_raycloud.ply",
  input_mode   = "raycloud",
  project_root = "D:/packageR",
  run_name     = "forest_resume_demo",
  use_colour   = FALSE,
  steps        = c("smooth", "info", "mesh"),
  resume       = TRUE,
  dry_run      = FALSE,
  echo         = TRUE
)
```

# Running individual step functions

Advanced users can run steps directly.

```{r eval = FALSE}
s2 <- ray_terrain(
  raycloud_file = "D:/packageR/data/plot/plot01_raycloud.ply",
  project_root  = "D:/packageR",
  out_dir       = "D:/packageR/runs/forest_stepwise/step4_terrain",
  dry_run       = FALSE,
  echo          = TRUE
)

s3 <- ray_trees(
  raycloud_file = "D:/packageR/data/plot/plot01_raycloud.ply",
  terrain_mesh  = s2$outputs$terrain_mesh,
  project_root  = "D:/packageR",
  out_dir       = "D:/packageR/runs/forest_stepwise/step5_trees",
  args_trees    = "--height_min 0.2 --distance_limit 3 --max_diameter 3 --gravity_factor 0.1",
  dry_run       = FALSE,
  echo          = TRUE
)

s4 <- ray_split_segmented(
  segmented_ply = s3$outputs$segmented_ply,
  project_root  = "D:/packageR",
  out_dir       = "D:/packageR/runs/forest_stepwise/step6_split",
  dry_run       = FALSE,
  echo          = TRUE
)
```

# Inspecting outputs

The forest wrapper returns a `ray_result` object with key paths such as:

- `raycloud_file`
- `terrain_mesh`
- `segmented_ply`
- `treefile`
- `split_segmented`
- `split_segmented_dir`
- `smoothed_treefile`
- `info_file`
- `mesh_file`

This makes it easier to continue from later stages or to integrate the workflow
into larger scripts.

# Notes

- For plot and forest workflows, starting from an existing raycloud is often
  the most stable route.
- Use `split_segmented_dir` to inspect the outputs of the split stage.
- Use `resume = TRUE` to continue later stages without re-running everything.
- Some raw imports may still produce warnings; fallback import modes can make
  the workflow more robust when an existing raycloud is available.
