---
title: "Getting started and single-tree workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started and single-tree workflow}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`rayforestr` provides a stage-based interface around RayCloudTools and TreeTools
through Docker. The package supports:

- step-level functions such as `ray_import()`, `ray_terrain()`, `ray_trees()`,
  `ray_smooth()`, `ray_info()`, and `ray_mesh()`
- high-level wrappers such as `ray_pipeline_single_tree()`
- resumable workflows
- multiple entry modes:
  - `point_cloud`
  - `raycloud`
  - `treefile`
  - `smoothed_treefile`

This vignette shows the recommended single-tree workflow.

# Prerequisites

Before running pipelines:

1. Install Docker and confirm that Docker is available from R.
2. Pull or build the required image.
3. Set the package image option.

```{r eval = FALSE}
library(rayforestr)

ray_check_docker()

options(
  rayforestr.raycloud_image = "ghcr.io/csiro-robotics/raycloudtools:latest"
)
```

In many setups, the same Docker image provides both RayCloudTools and TreeTools.
`rayforestr` will reuse the RayCloudTools image for TreeTools commands when no
separate TreeTools image is configured.

# Recommended project structure

A simple project layout is:

```text
project_root/
  data/
    single_tree/
      spruce_4.ply
      spruce_4_raycloud.ply
  runs/
```

Use a project root outside the package source folder.

# Single-tree workflow from an existing raycloud

This is often the most reliable starting point for single-tree processing.

```{r eval = FALSE}
library(rayforestr)

project_root <- "D:/packageR"
raycloud_file <- "D:/packageR/data/single_tree/spruce_4_raycloud.ply"

res <- ray_pipeline_single_tree(
  input_file   = raycloud_file,
  input_mode   = "raycloud",
  project_root = project_root,
  run_name     = "single_tree_from_raycloud",
  use_colour   = FALSE,
  steps        = c("terrain", "trees", "smooth", "info", "mesh"),
  resume       = FALSE,
  dry_run      = FALSE,
  echo         = TRUE
)

res
```

Expected outputs are written into stage folders such as:

- `step4_terrain`
- `step5_trees`
- `step7_smooth`
- `step8_info`
- `step9_mesh`

# Single-tree workflow from a raw point cloud

If you want to begin from a raw point cloud, use `input_mode = "point_cloud"`.

```{r eval = FALSE}
res <- ray_pipeline_single_tree(
  input_file   = "D:/packageR/data/single_tree/spruce_4.ply",
  input_mode   = "point_cloud",
  project_root = "D:/packageR",
  run_name     = "single_tree_from_point_cloud",
  use_colour   = FALSE,
  steps        = c("import", "terrain", "trees"),
  resume       = FALSE,
  dry_run      = FALSE,
  echo         = TRUE
)
```

## Import reuse modes

For difficult point-cloud inputs, `ray_import()` and the wrappers support three
practical modes:

- `reuse_existing = "never"`  
  Always attempt a fresh import.

- `reuse_existing = "prefer"`  
  Use an existing raycloud immediately.

- `reuse_existing = "fallback"`  
  Attempt a fresh import first, but fall back to an existing raycloud if needed.

### Example: explicit reuse of an existing raycloud

```{r eval = FALSE}
res_import <- ray_import(
  input_file = "D:/packageR/data/single_tree/spruce_4.ply",
  project_root = "D:/packageR",
  out_dir = "D:/packageR/runs/import_prefer/step2_rayimport",
  existing_raycloud = "D:/packageR/data/single_tree/spruce_4_raycloud.ply",
  reuse_existing = "prefer",
  dry_run = FALSE,
  echo = TRUE
)

res_import
```

### Example: wrapper fallback to a sibling raycloud

```{r eval = FALSE}
res <- ray_pipeline_single_tree(
  input_file = "D:/packageR/data/single_tree/spruce_4.ply",
  input_mode = "point_cloud",
  project_root = "D:/packageR",
  run_name = "single_tree_import_fallback",
  reuse_existing_import = "fallback",
  search_sibling_raycloud = TRUE,
  use_colour = FALSE,
  steps = c("import", "terrain", "trees"),
  resume = FALSE,
  dry_run = FALSE,
  echo = TRUE
)
```

In practice, RayCloudTools may sometimes print import warnings while still
creating a usable `*_raycloud.ply`. In other cases, `rayforestr` can continue by
reusing an existing sibling raycloud when fallback is enabled.

# Starting from an existing treefile

If you already have a treefile, you can continue with later steps only.

```{r eval = FALSE}
res <- ray_pipeline_single_tree(
  input_file   = "D:/packageR/runs/example/step5_trees/spruce_4_raycloud_trees.txt",
  input_mode   = "treefile",
  project_root = "D:/packageR",
  run_name     = "single_tree_from_treefile",
  steps        = c("smooth", "info", "mesh"),
  resume       = FALSE,
  dry_run      = FALSE,
  echo         = TRUE
)
```

# Starting from an existing smoothed treefile

```{r eval = FALSE}
res <- ray_pipeline_single_tree(
  input_file   = "D:/packageR/runs/example/step7_smooth/spruce_4_raycloud_trees_smoothed.txt",
  input_mode   = "smoothed_treefile",
  project_root = "D:/packageR",
  run_name     = "single_tree_from_smoothed",
  steps        = c("info", "mesh"),
  resume       = FALSE,
  dry_run      = FALSE,
  echo         = TRUE
)
```

# Running individual step functions

Advanced users can call the step functions directly.

```{r eval = FALSE}
s2 <- ray_terrain(
  raycloud_file = "D:/packageR/data/single_tree/spruce_4_raycloud.ply",
  project_root  = "D:/packageR",
  out_dir       = "D:/packageR/runs/stepwise_demo/step4_terrain",
  dry_run       = FALSE,
  echo          = TRUE
)

s3 <- ray_trees(
  raycloud_file = "D:/packageR/data/single_tree/spruce_4_raycloud.ply",
  terrain_mesh  = s2$outputs$terrain_mesh,
  project_root  = "D:/packageR",
  out_dir       = "D:/packageR/runs/stepwise_demo/step5_trees",
  args_trees    = "--height_min 0.2 --distance_limit 3 --max_diameter 3 --gravity_factor 0.1",
  dry_run       = FALSE,
  echo          = TRUE
)
```

# Resume and selective steps

The wrapper can run only selected steps and resume later.

```{r eval = FALSE}
res_part1 <- ray_pipeline_single_tree(
  input_file   = "D:/packageR/data/single_tree/spruce_4_raycloud.ply",
  input_mode   = "raycloud",
  project_root = "D:/packageR",
  run_name     = "single_tree_resume_demo",
  steps        = c("terrain", "trees"),
  resume       = FALSE,
  dry_run      = FALSE,
  echo         = TRUE
)

res_part2 <- ray_pipeline_single_tree(
  input_file   = "D:/packageR/data/single_tree/spruce_4_raycloud.ply",
  input_mode   = "raycloud",
  project_root = "D:/packageR",
  run_name     = "single_tree_resume_demo",
  steps        = c("smooth", "info", "mesh"),
  resume       = TRUE,
  dry_run      = FALSE,
  echo         = TRUE
)
```

# Interpreting outputs

`ray_pipeline_single_tree()` returns a `ray_result` object with key output paths,
for example:

- `raycloud_file`
- `terrain_mesh`
- `segmented_ply`
- `treefile`
- `smoothed_treefile`
- `info_file`
- `mesh_file`

This makes it easier to continue programmatically from later stages.

# Notes

- For single-tree work, starting from an existing raycloud is often the most
  reliable route.
- Some raw point clouds may trigger import warnings or degenerate imports.
- `rayforestr` supports practical recovery through existing-raycloud reuse and
  fallback workflows.
