---
title: "Import strategies and troubleshooting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Import strategies and troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`rayforestr` wraps `rayimport` and related RayCloudTools commands through Docker.
In practice, raw point-cloud import can behave in several ways:

- import succeeds normally
- import prints warnings but still creates a usable `*_raycloud.ply`
- import fails to create a usable raycloud
- import becomes practical only when an existing raycloud is reused

This vignette explains the import strategies available in `rayforestr` and how
to handle common issues.

# The three reuse modes

`ray_import()` supports three practical reuse strategies through
`reuse_existing`:

- `"never"`  
  Always attempt a fresh import.

- `"prefer"`  
  Use an existing raycloud immediately.

- `"fallback"`  
  Attempt a fresh import first, but fall back to an existing raycloud when
  needed.

The pipeline wrappers expose the same idea through
`reuse_existing_import`.

# Typical import patterns

## 1. Fresh import only

Use this when you want to force a new import.

```{r eval = FALSE}
res <- ray_import(
  input_file = "D:/packageR/data/single_tree/spruce_4.ply",
  project_root = "D:/packageR",
  out_dir = "D:/packageR/runs/import_never/step2_rayimport",
  reuse_existing = "never",
  dry_run = FALSE,
  echo = TRUE
)
```

## 2. Prefer an existing raycloud

Use this when you already trust an existing raycloud and do not want to rerun
raw import.

```{r eval = FALSE}
res <- ray_import(
  input_file = "D:/packageR/data/single_tree/spruce_4.ply",
  project_root = "D:/packageR",
  out_dir = "D:/packageR/runs/import_prefer/step2_rayimport",
  existing_raycloud = "D:/packageR/data/single_tree/spruce_4_raycloud.ply",
  reuse_existing = "prefer",
  dry_run = FALSE,
  echo = TRUE
)

res
```

## 3. Try import first, then fall back

Use this when you want a fresh import if possible, but still want the workflow
to continue if an existing raycloud is available.

```{r eval = FALSE}
res <- ray_import(
  input_file = "D:/packageR/data/single_tree/spruce_4.ply",
  project_root = "D:/packageR",
  out_dir = "D:/packageR/runs/import_fallback/step2_rayimport",
  reuse_existing = "fallback",
  search_sibling_raycloud = TRUE,
  dry_run = FALSE,
  echo = TRUE
)

res
```

# Wrapper-level import fallback

The same strategy can be used directly in the wrappers.

## Single-tree wrapper

```{r eval = FALSE}
res <- ray_pipeline_single_tree(
  input_file = "D:/packageR/data/single_tree/spruce_4.ply",
  input_mode = "point_cloud",
  project_root = "D:/packageR",
  run_name = "single_tree_import_fallback",
  reuse_existing_import = "fallback",
  search_sibling_raycloud = TRUE,
  use_colour = FALSE,
  steps = c("import", "terrain", "trees"),
  resume = FALSE,
  dry_run = FALSE,
  echo = TRUE
)
```

## Forest wrapper

```{r eval = FALSE}
res <- ray_pipeline_forest(
  input_file = "D:/packageR/data/plot/plot01.ply",
  input_mode = "point_cloud",
  project_root = "D:/packageR",
  run_name = "forest_import_fallback",
  reuse_existing_import = "fallback",
  search_sibling_raycloud = TRUE,
  use_colour = FALSE,
  steps = c("import", "terrain", "trees", "split"),
  resume = FALSE,
  dry_run = FALSE,
  echo = TRUE
)
```

# Understanding noisy imports

In some cases, RayCloudTools may print warnings such as:

- no times provided
- suspicious large coordinates
- zero-intensity non-returns

This does **not always mean** that no output raycloud was created.

A practical pattern observed in real workflows is:

1. `rayimport` prints warnings or an error message
2. a `*_raycloud.ply` is still written
3. downstream steps such as terrain and trees can still run successfully

Because of this, the most important criterion is not only the console message,
but whether a usable `*_raycloud.ply` exists after the step.

# Sibling raycloud discovery

When `search_sibling_raycloud = TRUE`, `ray_import()` can look for a raycloud
next to the input point cloud using the expected naming pattern, for example:

- input: `spruce_4.ply`
- sibling raycloud: `spruce_4_raycloud.ply`

This is especially useful when:

- the raw `.ply` is difficult to import
- a raycloud was created previously outside the current run
- you want reproducible recovery without manually editing the workflow

# Starting from a raycloud directly

If you already know a valid raycloud exists, the cleanest approach is often to
skip point-cloud import entirely and start from:

```{r eval = FALSE}
res <- ray_pipeline_single_tree(
  input_file = "D:/packageR/data/single_tree/spruce_4_raycloud.ply",
  input_mode = "raycloud",
  project_root = "D:/packageR",
  run_name = "single_tree_from_raycloud",
  use_colour = FALSE,
  steps = c("terrain", "trees", "smooth", "info", "mesh"),
  resume = FALSE,
  dry_run = FALSE,
  echo = TRUE
)
```

This is often the most reliable path for single-tree processing.

# Common problems and responses

## Problem: Docker is not available

Check Docker first:

```{r eval = FALSE}
ray_check_docker()
```

If this fails, fix Docker before running the pipeline.

## Problem: Import says it failed

Check whether a raycloud actually exists:

```{r eval = FALSE}
list.files("D:/packageR/runs/example/step2_rayimport", full.names = TRUE)
```

If a usable raycloud exists, continue from that file or use fallback mode.

## Problem: You already have a raycloud

Do not force a raw import unless you need to.

Use either:

- `input_mode = "raycloud"`
- `reuse_existing = "prefer"`

## Problem: You want recovery but not silent skipping

Use:

```{r eval = FALSE}
reuse_existing = "fallback"
```

This tries fresh import first, then reuses an existing raycloud only when needed.

## Problem: You want a completely fresh run

Use:

```{r eval = FALSE}
reuse_existing = "never"
```

and choose a fresh run directory.

# Recommended decision guide

A practical decision order is:

1. If you already trust a raycloud, start from `input_mode = "raycloud"`.
2. If you want to test a fresh import but still want robustness, use fallback.
3. If you want strict import behavior for debugging, use `reuse_existing = "never"`.

# Notes

- For difficult single-tree inputs, an existing raycloud is often the most
  reliable starting point.
- Console warnings during import do not automatically mean that no usable
  output was created.
- `rayforestr` is designed to support practical workflows by allowing reuse and
  fallback strategies instead of forcing unnecessary failure.
